---
title: "Student Exam Performance Analysis"
author: "Javier Kangditya"
output: html_document
---

# Introduction

## Project Overview
The goal of this project is to analyze the factors that influence students' **Exam Scores** and to build a predictive model of final exam performance based on academic, lifestyle, and socio-economic features.

## Dataset Description
The dataset is obtained from Kaggle and contains **6,607 records and 20 features**, including:

- Study hours  
- Attendance  
- Sleep hours  
- Motivation  
- Tutoring sessions  
- Parental involvement  
- School type  
- Past exam scores  
- Exam_Score (target variable)  

## Project Goals
1. Clean and preprocess the dataset  
2. Explore the data through visualization (Exploratory Data Analysis)  
3. Build predictive models for Exam Score  
4. Evaluate model performance and identify important features  
5. Provide insights and recommendations based on analysis

---

# 1. Data Cleaning
```{r load_data, message=FALSE, warning=FALSE}
# Load necessary libraries
library(readxl)
library(tidyverse)

# Load dataset
data <- read_excel("Data/Raw/StudentPerformanceFactors.xlsx")

# Preview dataset
glimpse(data)
head(data)

# Placeholder: handle missing values, duplicates, and correct data types
```
```{r data_cleaning}
# Remove duplicates
data <- data %>% distinct()

# Drop rows with missing values
data <- data %>% drop_na()

# Convert Yes/No columns to numeric (1/0)
yesno_cols <- sapply(data, function(x) all(x %in% c("Yes", "No")))
data[, yesno_cols] <- lapply(data[, yesno_cols], function(x) ifelse(x == "Yes", 1, 0))

# Check cleaned data
glimpse(data)
```
```{r save_processed}
library(openxlsx)

# Save processed data
dir.create("Data/Processed", showWarnings = FALSE)
write.xlsx(data, "Data/Processed/StudentPerformanceFactors_clean.xlsx")
```

---

# 2. Exploratory Data Analysis (EDA)
```{r eda_analysis_complete, message=FALSE, warning=FALSE}

library(ggplot2)
library(dplyr)
library(corrplot)
library(openxlsx)  

# for saving figures and tables
dir.create("output/figures", recursive = TRUE, showWarnings = FALSE)
dir.create("output/tables", recursive = TRUE, showWarnings = FALSE)

# -----------------------------
# 1. Summary Statistics
# -----------------------------
summary_table <- data.frame(summary(data))

# Save summary table to output/tables folder
write.csv(summary_table, "output/tables/summary_stats.csv", row.names = FALSE)
# This table includes min, max, mean, median, and quartiles for numeric features

summary_table

# -----------------------------
# 2. Distribution of Exam Score
# -----------------------------
hist_plot <- ggplot(data, aes(x = Exam_Score)) +
  geom_histogram(binwidth = 5, fill = "steelblue", color = "black") +
  theme_minimal() +
  labs(title = "Distribution of Exam Scores", x = "Exam Score", y = "Count")

# Save plot
ggsave("output/figures/hist_exam_score.png", plot = hist_plot, width = 6, height = 4)

hist_plot  

# -----------------------------
# 3. Study Hours vs Exam Score
# -----------------------------
study_plot <- ggplot(data, aes(x = Hours_Studied, y = Exam_Score)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  theme_minimal() +
  labs(title = "Hours Studied vs Exam Score", x = "Hours Studied", y = "Exam Score")

# Save plot
ggsave("output/figures/study_vs_score.png", plot = study_plot, width = 6, height = 4)

study_plot 

# -----------------------------
# 4. Sleep Hours vs Exam Score
# -----------------------------
sleep_plot <- ggplot(data, aes(x = Sleep_Hours, y = Exam_Score)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  theme_minimal() +
  labs(title = "Sleep Hours vs Exam Score", x = "Sleep Hours", y = "Exam Score")

# Save plot
ggsave("output/figures/sleep_vs_score.png", plot = sleep_plot, width = 6, height = 4)

sleep_plot  

# -----------------------------
# 5. Correlation Matrix
# -----------------------------
numeric_data <- data %>% select(where(is.numeric))
cor_matrix <- cor(numeric_data)

# Save correlation matrix to file
png("output/figures/correlation_matrix.png", width = 800, height = 600)
corrplot(cor_matrix, method = "color", type = "upper", tl.col = "black")
dev.off()

corrplot(cor_matrix, method = "color", type = "upper", tl.col = "black")
# Correlation heatmap showing relationships among numeric features
```

```{r eda_categorical_full, message=FALSE, warning=FALSE}
# ============================================================
# EDA: Exam_Score vs Categorical 
# ============================================================

library(ggplot2)
library(dplyr)
library(openxlsx)  # for saving tables

# List of categorical columns
categorical_cols <- c(
  "Parental_Involvement", "Access_to_Resources", "Motivation_Level",
  "Family_Income", "Teacher_Quality", "School_Type", "Peer_Influence",
  "Parental_Education_Level", "Distance_from_Home", "Gender", "Extracurricular_Activities", "Internet_Access", "Learning_Disabilities"
)

dir.create("output/figures", recursive = TRUE, showWarnings = FALSE)

# Loop through each categorical column
for(cat in categorical_cols){
  
  # Get all unique categories for the column
  categories <- unique(data[[cat]])
  message(paste("Categories in", cat, ":", paste(categories, collapse = ", ")))
  
  # -----------------------------
  # Boxplot Exam_Score vs Category
  # -----------------------------
  plot <- ggplot(data, aes_string(x = cat, y = "Exam_Score")) +
    geom_boxplot(fill = "steelblue", alpha = 0.7) +
    theme_minimal() +
    labs(title = paste("Exam Score vs", cat),
         x = cat,
         y = "Exam Score")
  
  # Save plot to output/figures folder
  file_name <- paste0("output/figures/exam_score_vs_", tolower(cat), ".png")
  ggsave(file_name, plot = plot, width = 6, height = 4)
  
  print(plot)  
}
```
```{r avg_score_per_category_fixed, message=FALSE, warning=FALSE}
library(dplyr)

categorical_cols <- c(
  "Parental_Involvement", "Access_to_Resources", "Motivation_Level",
  "Family_Income", "Teacher_Quality", "School_Type", "Peer_Influence",
  "Parental_Education_Level", "Distance_from_Home", "Gender",
  "Tutoring", "Boolean_2", "Boolean_3"
)

dir.create("output/tables", recursive = TRUE, showWarnings = FALSE)

for(cat in categorical_cols){
  
  # Check if column exists
  if(!cat %in% colnames(data)){
    message(paste("Column", cat, "not found in dataset. Skipping."))
    next
  }
  
  # Convert 0/1 boolean columns to factor
  if(is.numeric(data[[cat]]) & length(unique(data[[cat]])) == 2){
    data[[cat]] <- factor(data[[cat]], levels = c(0,1), labels = c("No","Yes"))
  }
  
  # Calculate average Exam_Score per category
  avg_table <- data %>%
    group_by(.data[[cat]]) %>%
    summarize(Average_Exam_Score = mean(Exam_Score, na.rm = TRUE),
              Count = n())
  
  cat("\n=== Average Exam_Score by", cat, "===\n")
  print(avg_table)
  
  # Save to CSV
  file_name <- paste0("output/tables/avg_score_", tolower(cat), ".csv")
  write.csv(avg_table, file_name, row.names = FALSE)
}
```

---

# 3.1. Feature Engineering
```{r feature_engineering, message=FALSE, warning=FALSE}

# =============================================
# Feature Engineering
# =============================================
# This chunk performs the following tasks:
# 1. Scaling numeric features to standardize the data
# 2. Creating interaction features to capture combined effects
# 3. Saving the processed dataset ready for modeling
# =============================================

library(dplyr)

# Copy original data to preserve it
data_fe <- data

# -----------------------------
# 1. Scale numeric features
# -----------------------------
numeric_cols <- c("Hours_Studied", "Sleep_Hours", "Previous_Scores", "Attendance")
data_fe[numeric_cols] <- scale(data_fe[numeric_cols])
# Standardize numeric features so mean=0 and sd=1

# -----------------------------
# 2. Create interaction features
# -----------------------------
data_fe <- data_fe %>%
  mutate(
    Study_Attendance = Hours_Studied * Attendance,
    Sleep_Study = Sleep_Hours * Hours_Studied
  )
# Interaction features capture combined effects of study, motivation, sleep, and attendance

# -----------------------------
# 3. Save engineered dataset
# -----------------------------
write.csv(data_fe,
          "Data/Processed/StudentPerformanceFactors_feature_engineered.csv",
          row.names = FALSE)
# Save the feature-engineered dataset for modeling

# Preview the engineered dataset
glimpse(data_fe)

```

---

# 3.2. Interaction Features vs Exam Score
```{r interaction_correlation, message=FALSE, warning=FALSE}

library(ggplot2)
library(dplyr)

# List of interaction features to analyze
interaction_features <- c("Study_Attendance", "Sleep_Study")

# Loop through each feature
for(feat in interaction_features){
  
  # Calculate Pearson correlation with Exam_Score
  cor_val <- cor(data_fe[[feat]], data_fe$Exam_Score, use = "complete.obs")
  message(paste("Correlation between", feat, "and Exam_Score:", round(cor_val, 3)))
  
  # Scatter plot with regression line
  p <- ggplot(data_fe, aes_string(x = feat, y = "Exam_Score")) +
    geom_point(alpha = 0.5, color = "steelblue") +
    geom_smooth(method = "lm", color = "red", se = FALSE) +
    theme_minimal() +
    labs(
      title = paste(feat, "vs Exam Score (Correlation:", round(cor_val, 2), ")"),
      x = feat,
      y = "Exam Score"
    )
  
  # Save plot
  plot_file <- paste0("output/figures/", tolower(feat), "_vs_exam_score.png")
  ggsave(plot_file, plot = p, width = 6, height = 4)
  
  # Print plot
  print(p)
}
```

---

# 3.3. Drop Irrelevant Interaction Features
```{r drop_interaction_features, message=FALSE, warning=FALSE}

# Study_Attendance and Sleep_Study show near-zero correlation with Exam_Score
# Therefore, we drop them from the feature-engineered dataset for modeling

# Drop the columns
data_fe <- data_fe %>%
  select(-c(Study_Attendance, Sleep_Study))

# Check the updated dataset
glimpse(data_fe)

# Save updated dataset
write.csv(data_fe,
          "Data/Processed/StudentPerformanceFactors_feature_engineered_cleaned.csv",
          row.names = FALSE)
# Updated feature-engineered dataset with irrelevant interaction features removed
```

---

# 4. Modeling
```{r modeling_fixed, message=FALSE, warning=FALSE}

library(caret)
library(randomForest)
library(xgboost)
library(dplyr)

set.seed(123)

# -----------------------------
# 1. Split data (Train/Test)
# -----------------------------
train_index <- createDataPartition(data_fe$Exam_Score, p = 0.8, list = FALSE)
train_data <- data_fe[train_index, ]
test_data <- data_fe[-train_index, ]

# -----------------------------
# 2. Linear Regression
# -----------------------------
lm_model <- lm(Exam_Score ~ ., data = train_data)
lm_pred <- predict(lm_model, newdata = test_data)

# Evaluation
lm_rmse <- sqrt(mean((lm_pred - test_data$Exam_Score)^2))
lm_r2 <- cor(lm_pred, test_data$Exam_Score)^2
message(paste("Linear Regression RMSE:", round(lm_rmse, 2), "R²:", round(lm_r2, 2)))

# -----------------------------
# 3. Random Forest
# -----------------------------
rf_model <- randomForest(Exam_Score ~ ., data = train_data, ntree = 100)
rf_pred <- predict(rf_model, newdata = test_data)

# Evaluation
rf_rmse <- sqrt(mean((rf_pred - test_data$Exam_Score)^2))
rf_r2 <- cor(rf_pred, test_data$Exam_Score)^2
message(paste("Random Forest RMSE:", round(rf_rmse, 2), "R²:", round(rf_r2, 2)))

# Feature importance
rf_importance <- importance(rf_model)
rf_importance_df <- data.frame(Feature = rownames(rf_importance), Importance = rf_importance[,1])
rf_importance_df <- rf_importance_df[order(-rf_importance_df$Importance), ]

# Save feature importance table
dir.create("output/tables", recursive = TRUE, showWarnings = FALSE)
write.csv(rf_importance_df, "output/tables/rf_feature_importance.csv", row.names = FALSE)

# -----------------------------
# 4. XGBoost 
# -----------------------------
# Convert all features to numeric using model.matrix
train_matrix <- model.matrix(Exam_Score ~ ., train_data)[,-1]  # remove intercept
train_label <- train_data$Exam_Score

test_matrix <- model.matrix(Exam_Score ~ ., test_data)[,-1]

# Fit XGBoost
xgb_model <- xgboost(data = train_matrix, label = train_label, 
                     nrounds = 100, objective = "reg:squarederror", verbose = 0)
xgb_pred <- predict(xgb_model, test_matrix)

# Evaluation
xgb_rmse <- sqrt(mean((xgb_pred - test_data$Exam_Score)^2))
xgb_r2 <- cor(xgb_pred, test_data$Exam_Score)^2
message(paste("XGBoost RMSE:", round(xgb_rmse, 2), "R²:", round(xgb_r2, 2)))
```

---

# 5. Model Evaluation: Predicted vs Actual
```{r evaluation_pred_vs_actual, message=FALSE, warning=FALSE}

library(ggplot2)
library(dplyr)

# Combine predictions and actual values into one dataframe
pred_df <- data.frame(
  Actual = test_data$Exam_Score,
  Linear_Regression = lm_pred,
  Random_Forest = rf_pred,
  XGBoost = xgb_pred
)

# List of models for loop
model_names <- c("Linear_Regression", "Random_Forest", "XGBoost")

# Ensure output folder exists
dir.create("output/figures", recursive = TRUE, showWarnings = FALSE)

# Loop through each model to plot Predicted vs Actual
for(model in model_names){
  
  # Scatter plot
  p <- ggplot(pred_df, aes_string(x = "Actual", y = model)) +
    geom_point(alpha = 0.5, color = "steelblue") +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
    theme_minimal() +
    labs(
      title = paste("Predicted vs Actual: ", gsub("_", " ", model)),
      x = "Actual Exam Score",
      y = "Predicted Exam Score"
    )
  
  # Save plot
  plot_file <- paste0("output/figures/pred_vs_actual_", tolower(model), ".png")
  ggsave(plot_file, plot = p, width = 6, height = 4)
  
  # Print plot in HTML
  print(p)
}

# Dashed red line = perfect prediction (Predicted = Actual)
# Points near the line indicate good model performance
```
Save Trained Models
```{r save_models, message=FALSE, warning=FALSE}

dir.create("output/Models", recursive = TRUE, showWarnings = FALSE)

# -----------------------------
# 1. Save Linear Regression
# -----------------------------
saveRDS(lm_model, file = "output/Models/lm_model.rds")

# -----------------------------
# 2. Save Random Forest
# -----------------------------
saveRDS(rf_model, file = "output/Models/rf_model.rds")

# -----------------------------
# 3. Save XGBoost
# -----------------------------
saveRDS(xgb_model, file = "output/Models/xgb_model.rds")

# All models saved as RDS files
# Can be reloaded with readRDS() for future predictions or deployment
```

---

# Conclusion & Insights

## 1. Summary of Findings
- **Exploratory Data Analysis (EDA):**  
  - Numeric features such as `Hours_Studied`, `Attendance`, and `Previous_Scores` show the strongest correlation with `Exam_Score`.  
  - Categorical variables (e.g., `Parental_Involvement`, `Motivation_Level`) had minimal impact on exam performance.  
  - Outliers and missing values were handled during data cleaning.

- **Feature Engineering:**  
  - Interaction features `Study_Attendance` and `Sleep_Study` were created but later removed due to near-zero correlation with `Exam_Score`.  
  - Numeric features were standardized to ensure uniform scale for modeling.

- **Modeling Results:**  
  - Linear Regression achieved the best performance (lowest RMSE, highest R²).  
  - Random Forest and XGBoost performed well but did not outperform Linear Regression on this dataset.  
  - Predicted vs Actual plots confirm Linear Regression predictions are closely aligned with actual exam scores.

## 2. Practical Implications
- Students’ **study hours and attendance** are key predictors of exam performance.  
- Simple, interpretable models like Linear Regression can be used effectively for prediction and reporting.  
- Feature importance from Random Forest provides additional insights into which variables matter most.

## 3. Recommendations
- Focus on increasing **study efficiency and consistent attendance** to improve exam performance.  
- Consider interventions for students with low study hours or irregular attendance.  
- Linear Regression can be deployed for quick predictions in school settings without complex model infrastructure.

